Type-checking Linearity in GHC Core

### 2024-01-24 icfp.tex
revised intro

Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
to existing type systems by enforcing that certain \emph{resources} (e.g.~a
file handle) must be used \emph{exactly once}.
%
In programming languages with a linear type system, not using certain resources
or using them twice is flagged as a type error. Linear types can thus be used
to, for instance, statically guarantee that socket descriptors are closed or
heap-allocated memory is freed, exactly once (leaks and double-frees become
type errors), or guarantee channel-based communication protocols are
deadlock-free~\cite{10.1007/978-3-642-15375-4_16},
% implement safe
% high-performance language interoperability~\cite{}, 
%guarantee that quantum entangled states are not duplicated~\cite{}
among other correctness properties~\cite{10.1145/3373718.3394765,10.1145/3527313,cite:linearhaskell}.
% handle mutable state safely~\cite{}

Consider the following program in which allocated memory is freed twice.
% We know this to be the dreaded double-free error which will crash the program at runtime.
Regardless of the \emph{double-free} error, a C-like type system will
accept such a program:
% \centering
\begin{code}
let p = malloc(4); in free(p); free(p);
\end{code}
Under the lens of a linear type system, if $p$ is deemed to be a
linear resource created by the call to \texttt{malloc}, it
must be used \emph{exactly once}.  However, the program above uses $p$ twice,
in the two calls to \texttt{free} and so fails to type-check.
%
In this sense, linear typing
effectively ensures the program does not compile with a double-free error.

Despite their promise and extensive presence in research
literature~\cite{Wadler1990LinearTC,CERVESATO2000133,10.1093/logcom/2.3.297},
an effective design combining linear and non-linear typing is both
challenging and necessary to bring the advantages of linear typing to
mainstream languages.
%
Consequently, few general purpose programming languages have linear
type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
% a linearly and dependently typed language based on Quantitative Type Theory,
Rust~\cite{10.1145/2692956.2663188}, a language whose
ownership types build on linear types to guarantee memory safety
without garbage collection or reference counting, and, more recently,
Haskell~\cite{cite:linearhaskell}, a pure, functional, and
\emph{lazy} general purpose programming language.
%
% Besides Haskell's supporting linear
% types according to the said paper, Idris 2\cite{} supports linear types in a
% dependently typed setting, Clean\cite{} has uniqueness types which are closely
% related to linear types, and Rust\cite{} has ownership types which build from
% linear types. 
%
Linearity in Haskell stands out from linearity in other languages:
%\begin{itemize}
    % \item Linear types were only added to the language roughly \emph{31 years
    %     after} Haskell's inception, unlike Rust and Idris 2 which were designed
    %     with linearity from the start. It is an especially difficult endeavour
    %     to add linear types to a well-established language with a large library
    %     ecosystem and many active users, rather than to develop the language
    %     from the ground up with linear types in mind. The successful
    %     approach as implemented in the Glasgow Haskell Compiler (GHC), the leading
    %     Haskell compiler, is based on Linear Haskell~\cite{cite:linearhaskell}, which describes a linear type system designed
    %     with retaining backwards-compatibility and allowing code reuse across
    %     linear and non-linear users of the same libraries.
    %     We describe Linear Haskell in detail in Section~\ref{sec:linear-haskell}.
%
%\item
  Linear types permeate Haskell down to (its) Core, the
        intermediate language into which Haskell is
        translated. Core is a minimal, explicitly typed,
        functional language, to which multiple
        Core-to-Core optimising transformations are
        applied during compilation. Due to Core's minimal design, typechecking
        Core programs is very efficient and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimised
        Core program fails to typecheck, the optimising
        transformations are very likely to have introduced an error
        in the resulting program.
        %We present Core and its formal
        %basis, System~$F_C$~\cite{cite:systemfc}, in Section~\ref{sec:core}.
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?
%
        % \item

        Both Haskell and its intermediate language Core are \emph{lazily}
        evaluated, i.e., expressions in Haskell are only evaluated when needed,
        unlike C or Rust, which are \emph{eagerly} evaluated.
        %
        Laziness allows an optimising compiler to aggressively transform the source
        program without changing its semantics and, indeed, the Glasgow Haskell
        Compiler (GHC) heavily transforms Core by leveraging its laziness.
        %
        However, lazy evaluation interacts non-trivially with linearity.
        Intuitively, since expressions are not necessarily evaluated, an
        occurrence of a linear resource in an expression does not necessarily
        entail consuming that resource (i.e., if the expression is not evaluated,
        the resource is not used).
%
        In eagerly evaluated languages, the distinction between syntactic uses of a
        resource and the actual use of linear resources at runtime does not exist --
        an occurrence of a variable in the program always entails consuming it.
        %
        This interaction is unique to Haskell since, as far as we know, it is the
        only language featuring both laziness and linearity.
        %
        %We review lazy and eager evaluation strategies in Section~\ref{sec:background:evaluation-strategies}.

%\end{itemize}
%

% Linear core good Aligned with the philosophy of having a \emph{typed}
% intermediate language, the integration of linear types in Haskell required
% extending Core with linear types.
%
Much like a typed Core ensures that the translation from Haskell (dubbed
\emph{desugaring}) and the subsequent optimising transformations applied to
Core are correctly implemented, a \emph{linearly typed} Core guarantees that
linear resource usage in the source language is not violated by the translation
process and the compiler optimisation passes.
%
It is crucial that a program's behaviour enforced by linear types is \emph{not}
changed by the compiler in the desugaring or optimisation stages (optimisations
should not destroy linearity!) and a linearity aware Core type-checker is key in
providing such guarantees -- it would be disastrous if the compiler, e.g.,
duplicated a pointer to heap-allocated memory that was previously just used
once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
Even more, linearity in Core can inform Core-to-Core optimising
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
including inlining and $\beta$-reduction, to produce more performant programs.

@@

In spite of a linearly typed Core ultimately being the desired intermediate
language for the Glasgow Haskell Compiler,
both in its expressiveness to completely represent a Haskell with
linear types and in enabling more extensive compiler validations plus improved optimisations,
linearity is effectively ignored in Core. The reason is not evidently clear:
if we can typecheck linearity in the surface level Haskell, it should also be
possible, and natural, to do so in Core.
% why do we fail to do so in Core?

The desugaring process from surface level Haskell to Core, and the subsequent
Core-to-Core optimising transformations, eliminate, and rearrange, most of the
syntactic constructs through which linearity checking is performed -- often
resulting in programs completely different from the original, especially due to
the flexibility laziness provides a compiler in the optimisations it may
perform. % without changing the program semantics.
%
Crucially, since optimisations do not destroy linearity, the resulting program
is still linear \emph{semantically}, however, the current linear type system
fails to recognize it as linear.
%
% The optimisations transform programs that are linear into ones that don't
% \emph{look} linear, but remain linear \emph{semantically}.
%
For instance, let $x$ be a linear resource in the two following programs, where
the latter results from inlining $y$ in the let body of the former. Despite the
result no longer \emph{looking} linear (as there are now two syntactic
occurrences of the linear resource $x$), the program \emph{is} indeed linear
because the let-bound expression freeing $x$ is never evaluated, so $x$ is
consumed exactly once when it is freed in the let body:
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

@@

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the call-by-need evaluation model
of Core, and thus a whole class of programs that are linear under the lens of
lazy evaluation are rejected by Core's current linear type system.

In this work, we address this limitation (and its implications on validating
and influencing optimising transformations) by designing a type system which
understands semantic linearity in the presence of laziness and is suitable for
the intermediate language of an optimising compiler.
%
In detail, our contributions are:
%
\begin{itemize}

\item We explain and provide insights into \emph{semantic} linearity in
contrast to \emph{syntactic} linearity, in Haskell, by example
($\S$~\ref{sec:linearity-semantically}).

\item We introduce Linear Core, a type system for a linear lazy language with
all key features from Core (except for type equality coercions), which,
crucially, understands semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to understand linearity
semantically in the context of lazy evaluation
($\S$~\ref{sec:main:linear-core}).

\item We prove Linear Core to be sound (well-typed Linear Core programs do not
get \emph{stuck}) and prove that multiple optimising transformations (which
currently violate linearity in Core) preserve types and linearity in Linear
Core ($\S$~\ref{sec:main:metatheory}).

\item We implement Linear Core as a GHC plugin which typechecks linearity in
all intermediate Core programs produced during the compilation process, showing
it accepts the intermediate programs resulting from (laziness-abusing)
transformations in linearity-heavy Haskell libraries, such as
\texttt{linear-base} ($\S$~\ref{sec:discuss:implementation}).

@@

A linear type system statically guarantees that linear resources are
\emph{consumed} exactly once. Consequently, whether a program is well-typed
under a linear type system intrinsically depends on the precise definition of
\emph{consuming} a resource. Even though \emph{consuming} a resource is
commonly regarded as synonymous with the \emph{syntactic occurrence} of the resource
in the program, that is not always the case.
%
In fact, this chapter highlights the distinction between using resources
\emph{syntactically} and \emph{semantically} as a principal limitation of
linear type systems for non-strict languages, with examples motivated by the
constructs available in GHC Core and how they are evaluated.

% and we show how a linear type system found on
% \emph{consuming resources semantically} rather than \emph{using resources
% syntactically} can accept

Consider the following program in a functional Haskell-like language,
where a computation that closes the given \lstinline{handle} is bound to \lstinline{x} before the
\lstinline{handle} is returned:
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%
In this example, the \lstinline{handle} appears to be closed before
returned, whereas in fact the handle will only be closed if the let bound
computation is effectively run (i.e. evaluated).
%
The example illustrates that \emph{consuming} a resource is not necessarily
synonymous with its syntactic occurrences, since depending on the evaluation strategy
of the language the computation that closes the handle might not be
executed, and if so, the \lstinline{handle} in that unused computation
is not consumed.
%
Expanding on this, consider the above example program under distinct
evaluation strategies: under call-by-value, %With \emph{eager}
% evaluation semantics,
the let
  bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
returned. It is clear that a linear type system should not accept such a
program since the linear resource \lstinline{handle} is duplicated -- it is used in a
computation that closes it, while still being made available to the caller of
the function;
%
on the other hand, with \emph{call-by-need}, the let
bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
used outside the scope of the function, so the handle is not closed by \lstinline{f}.
Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
let binding only results in the resource being \emph{consumed} if the binding
itself is \emph{consumed}. We argue that a linear type system under
\emph{call-by-need} evaluation should accept the above program, unlike a linear
type system for the same program evaluated \emph{call-by-value}.

>>>>

Linear type systems~\cite{cite:linear-logic,cite:barberdill} add expressiveness
to existing type systems by enforcing that certain \emph{resources} (e.g.~a
file handle) must be used \emph{exactly once}.
%
In programming languages with a linear type system, not using certain resources
or using them twice is flagged as a type error. Linear types can thus be used
to, for instance, statically guarantee that socket descriptors are closed or
heap-allocated memory is freed, exactly once (leaks and double-frees become
type errors), or guarantee channel-based communication protocols are
deadlock-free~\cite{10.1007/978-3-642-15375-4_16},
% implement safe
% high-performance language interoperability~\cite{}, 
%guarantee that quantum entangled states are not duplicated~\cite{}
among other correctness properties~\cite{10.1145/3373718.3394765,10.1145/3527313,cite:linearhaskell}.
% handle mutable state safely~\cite{}
%
For instance, consider the following program in which allocated memory is freed twice.
% We know this to be the dreaded double-free error which will crash the program at runtime.
Regardless of the \emph{double-free} error, a C-like type system will
accept such a program:
% \centering
\begin{code}
let p = malloc(4); in free(p); free(p);
\end{code}
Under the lens of a linear type system, if $p$ is deemed to be a
linear resource created by the call to \texttt{malloc}, it
must be used \emph{exactly once}.  However, the program above uses $p$ twice,
in the two calls to \texttt{free} and so fails to type-check.
%
In this sense, linear typing
effectively ensures the program does not compile with a double-free error.

Despite their promise and extensive presence in research
literature~\cite{Wadler1990LinearTC,CERVESATO2000133,10.1093/logcom/2.3.297},
an effective design combining linear and non-linear typing is both
challenging and necessary to bring the advantages of linear typing to
mainstream languages.
%
Consequently, few general purpose programming languages have linear
type systems. Among them are Idris~2~\cite{brady:LIPIcs.ECOOP.2021.9},
% a linearly and dependently typed language based on Quantitative Type Theory,
Rust~\cite{10.1145/2692956.2663188}, a language whose
ownership types build on linear types to guarantee memory safety
without garbage collection or reference counting, and, more recently,
Haskell~\cite{cite:linearhaskell}, a pure, functional, and
\emph{lazy} general purpose programming language.
%
% Besides Haskell's supporting linear
% types according to the said paper, Idris 2\cite{} supports linear types in a
% dependently typed setting, Clean\cite{} has uniqueness types which are closely
% related to linear types, and Rust\cite{} has ownership types which build from
% linear types. 
%

However, linearity in Haskell stands out from linearity in other
languages given that linear types permeate Haskell down to (its) Core,
the intermediate language into which Haskell is translated. Core is a
minimal, explicitly typed, functional language, to which multiple
Core-to-Core optimising transformations are applied during
compilation. Due to Core's minimal design, typechecking Core programs
is very efficient and doing so serves as a sanity check to the
correction of the source transformations. If the resulting optimised
Core program fails to typecheck, the optimising transformations are
very likely to have introduced an error in the resulting program.
%

Both Haskell and its intermediate language Core are \emph{lazily}
        evaluated, i.e., expressions in Haskell are only evaluated when needed,
        (unlike Rust, which is \emph{eagerly} evaluated).
        %
        % Laziness allows an optimising compiler to aggressively transform the source
        % program without changing its semantics and, indeed, the Glasgow Haskell
        % Compiler (GHC) heavily transforms Core by leveraging its laziness.
        %
        As we make clear in this paper, lazy evaluation interacts non-trivially with linearity.
        Intuitively, since expressions are not necessarily evaluated,
        a syntactic occurrence of a linear resource in an expression does not necessarily
        entail consuming that resource (i.e., if the expression is not evaluated,
        the resource is not used).
%
        In eagerly evaluated languages, the distinction between syntactic uses of a
        resource and the actual use of linear resources at runtime does not exist --
        an occurrence of a variable in the program always entails consuming it.
        %
        This interaction is unique to Haskell since, to the best of
        our knowledge, it is the
        only language featuring both laziness and linearity.

Just as Core's type system provides a degree of validation to the
translation from Haskell (dubbed \emph{desugaring}) and the subsequent
optimising transformations, a \emph{linearly typed} Core would guarantee that
linear resource usage in the source language is not violated by desugaring
and the compiler optimisation passes. Moreover, linearity information in
Core can inform Core-to-Core optimising 
transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
such as inlining and $\beta$-reduction, to produce more efficient programs.
%
% It is crucial that a program's behaviour enforced by linear types is \emph{not}
% changed by the compiler in the desugaring or optimisation stages (optimisations
% should not destroy linearity!) and a linearity aware Core type-checker is key in
% providing such guarantees -- it would be disastrous if the compiler, e.g.,
% duplicated a pointer to heap-allocated memory that was previously just used
% once and then freed in the original program.
%TODO: linearidade pode informar otimizacoes
%
% Even more, linearity in Core can inform Core-to-Core optimising
% transformations~\cite{cite:let-floating,peytonjones1997a,cite:linearhaskell},
% including inlining and $\beta$-reduction, to produce more performant programs.

@@

Despite of the advantages of a linearly typed Core, the status quo is
that linearity is effectively ignored in Core, despite being checked
in the source Haskell code. The reason is that the desugaring
and the subsequent Core-to-Core optimising transformations,
eliminate and rearrange most of the syntactic constructs through
which linearity checking is performed -- often resulting in programs
that are syntactically very different from the original, especially
due to the flexibility laziness provides a compiler in the
optimisations it may perform.
%
Crucially, while optimisations aim to preserve the semantics of
programs and so preserve a program's linear resource consumption,
a \emph{naive} syntactic analysis of the optimised programs (as
performed by the current Core type checker) often fails to recognize
the program as linear.
%
For instance, let $x$ be a linear resource in the two following programs, where
the latter results from inlining $y$ in the let body of the former. Despite the
the program no longer appearing linear (as there are now two 
occurrences of $x$), the program \emph{is} indeed linear
because the let-bound expression is never evaluated, so $x$ is
consumed exactly once (when it is freed in the let body):
% executed, $x$ will be freed exactly once:
%
% , \emph{is} linear, as the let bound expression freeing $x$
% is never evaluated (because is not needed) -- thus $x$ is only consumed once
% when freed in the let body:
\[
\begin{array}{ccc}
\llet{y = \textsf{free}~x}{y} & \Longrightarrow_{Inlining} & \llet{y = \textsf{free}~x}{\textsf{free}~x}
\end{array}
\]

@@

The Core optimising transformations expose a fundamental limitation of Core's
linear type system: it does not account for the call-by-need evaluation model
of Core, and thus a whole class of programs that are linear under the lens of
lazy evaluation are rejected by Core's current linear type system.
%
In this work, we address this limitation (and its implications on validating
and influencing optimising transformations) by designing a type system which
understands \emph{semantic linearity} in the presence of laziness and is suitable for
the intermediate language of an optimising compiler such as the
Glasgow Haskell Compiler.
%
Our contributions are as follows:
%
\begin{itemize}

\item We introduce the concept of \emph{semantic} linearity, in
contrast to \emph{syntactic} linearity in Haskell, by example
($\S$~\ref{sec:linearity-semantically}).

\item We introduce Linear Core, a type system for a linear lazy language with
the key features from Core (except for type equality coercions), which,
accounts for semantic linearity in the presence of laziness. To the
best of our knowledge, this is the first type system to capture linearity
semantically in this context
($\S$~\ref{sec:main:linear-core}).

\item We show the type soundness of Linear Core and prove that
  multiple optimising transformations, which
fail to type check in Core, are validated by Linear
Core ($\S$~\ref{sec:main:metatheory}).

\item We implement Linear Core as a GHC plugin which checks linearity in
all intermediate Core programs produced during the compilation process, showing
it accepts the programs resulting from transformations in libraries such as
\texttt{linear-base} ($\S$~\ref{sec:discuss:implementation}).

@@

A linear type system statically guarantees that linear resources are
\emph{consumed} exactly once. Consequently, whether a program is
well-typed under a linear type system intrinsically depends on the
definition of \emph{consuming} a resource.
In general, \emph{consuming} a resource is equated with its 
\emph{syntactic occurrence} in the program. However, as this section
makes clear, when considering linear type systems for non-strict
languages, a distinction between using
resources \emph{syntactically} and \emph{semantically} is needed.
%
Consider the following program in a functional Haskell-like language,
where a computation that closes the given \lstinline{handle} is bound to \lstinline{x} before the
\lstinline{handle} is returned:
%
\begin{lstlisting}
f : Handle *' $\lolli$ '* Handle
f handle = let x = close handle in handle
\end{lstlisting}
%

In the code above, the \lstinline{handle} appears to be closed before
returned, whereas in fact the handle will only be closed if the let bound
computation is effectively run (i.e.,~evaluated).
%
The example illustrates that \emph{consuming} a resource is not necessarily
synonymous with its syntactic occurrences, since depending on the evaluation strategy
of the language the computation that closes the handle might not be
executed, and if so, the \lstinline{handle} in that unused computation
is not consumed.
%
Expanding on this, consider the above example program under distinct
evaluation strategies: under call-by-value, %With \emph{eager}
% evaluation semantics,
the let
  bound expression \lstinline{close handle} is eagerly evaluated, and the \lstinline{handle} will be closed before being
returned. It is clear that a linear type system should not accept such a
program since the linear resource \lstinline{handle} is duplicated -- it is used in a
computation that closes it, while still being made available to the caller of
the function;
%
on the other hand, with \emph{call-by-need}, the let
bound expression will only be evaluated when the binding \lstinline{x} is needed. We return
the \lstinline{handle} right away, and the let binding is forgotten as it cannot be
used outside the scope of the function, so the handle is not closed by \lstinline{f}.
Under the lens of \emph{call-by-need} evaluation, \emph{using} a resource in a
let binding only results in the resource being \emph{consumed} if the binding
itself is \emph{consumed}. We argue that a linear type system under
\emph{call-by-need} evaluation should accept the above program, unlike a linear
type system for the same program evaluated \emph{call-by-value}.
